To be FAIR: Theory Specification Needs an Update

### 2025-06-06 manuscript.Rmd
add first draft for connection to prereg

With regard to the state of Interoperability in psychology, 
Kurt Lewin's [-@lewinPsychologyProcessGroup1943] adage "there's nothing as practical as a good theory" paints a hopeful picture of theories as useful tools in psychological researchers' day-to-day work.
But, as we argued, this is not the case.
The examples of X-Interoperability offered in Table \@ref(tab:tabmeehl) illustrate that much can be gained by integrating theory directly into analysis workflows, and by making theory X-Interoperable within software used for analysis.
For example, Interoperable theory could be used
to select control variables for causal inference [@cinelliCrashCourseGood2022],
or to preregister the inferential procedure that would lead to specific modifications of a theory after analyzing empirical data [@peikertReproducibleResearchTutorial2021],
or to derive machine-readable hypotheses [@lakensImprovingTransparencyFalsifiability2021] which could be automatically evaluated through integration testing [@vanlissaUsingEndpointsCheck2023].
Furthermore, theories can be X-Interoperable with each other to enable nesting, or using one theory to clarify elements of another theory.
For example, it should be possible to embed a theory about emotion regulation [e.g., @grossEmotionRegulationCurrent2015] within a theory of emotion regulation development [@morrisRoleFamilyContext2007].

>>>>

With regard to the state of Interoperability in psychology, 
Kurt Lewin's [-@lewinPsychologyProcessGroup1943] adage "there's nothing as practical as a good theory" paints a hopeful picture of theories as useful tools in psychological researchers' day-to-day work.
But, as we argued, this is not the case.
The examples of X-Interoperability offered in Table \@ref(tab:tabmeehl) illustrate that much can be gained by integrating theory directly into analysis workflows, and by making theory X-Interoperable within software used for analysis.
For example, Interoperable theory could be used
to select control variables for causal inference [@cinelliCrashCourseGood2022],
or to derive machine-readable hypotheses [@lakensImprovingTransparencyFalsifiability2021] which could be automatically evaluated through integration testing [@vanlissaUsingEndpointsCheck2023].
Furthermore, theories can be X-Interoperable with each other to enable nesting, or using one theory to clarify elements of another theory.
For example, it should be possible to embed a theory about emotion regulation [e.g., @grossEmotionRegulationCurrent2015] within a theory of emotion regulation development [@morrisRoleFamilyContext2007].

An important aspect of Interoperability lies in linking FAIR theories with existing open science methodologies---most notably, preregistration.
As noted, preregistration alone has not resolved the replication crisis in the social sciences.
This raises a fundamental question: What additional value does a FAIR theory provide, and how does it complement preregistration?

One distinction is quantitative:
FAIR theories operate at a higher level of abstraction.
They articulate general principles and expectations that cut across studies, whereas preregistrations focus on specific applications, analytic strategies, and study designs.
More fundamentally, however, there is a qualitative difference.
A FAIR theory is designed to make predictions and be revised in light of empirical results; preregistration, by contrast, prescribes how a particular study will be conducted, regardless of what the outcomes turn out to be.

This difference can be illustrated with a Bayesian analogy.
A preregistration is akin to a statistical model---it formalizes assumptions about data generation but is in itself indifferent to the data.
Combining a theory with a specific sampling strategy, study design, measurement, etc. allows one to make predictions even before observing data.
However, after observing data, the theory might require revision.
In this sense, a theory is first a prior, and then after conducting a specific investigation as prescribed by the preregistration, is refined to become a posterior.
In short, a preregistration rooted in a FAIR theory not only supports transparency, but also strengthens the epistemic cycle of prediction, testing, and revision---moving us closer to a cumulative science.
